
# Practical Machine Learning Project

## Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

## Objectives

You should create a report describing how you built your model

How you used cross validation

What you think the expected out of sample error is

Why you made the choices you did

You will also use your prediction model to predict 20 different test cases

Please constrain the text of the writeup to < 2000 words and the number of figures to be less than 5

link to a Github repo with your R markdown and compiled HTML file describing your analysis

## Data

Six young health participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions: exactly according to the specification (Class A), throwing the elbows to the front (Class B), lifting the dumbbell only halfway (Class C), lowering the dumbbell only halfway (Class D) and throwing the hips to the front (Class E).

Class A corresponds to the specified execution of the exercise, while the other 4 classes correspond to common mistakes.

The training data for this project are available here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

```{r}
# get the training and test data
if (!file.exists("pml-training.csv"))
{
    download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv",destfile="pml-training.csv",method="libcurl")
}
if (!file.exists("pml-testing.csv"))
{
    download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv",destfile="pml-testing.csv",method="libcurl")
}

# load the training and test data sets
training = read.csv("pml-training.csv", na.strings = c("", "NA", "#DIV/0!"))
testing = read.csv("pml-testing.csv", na.strings = c("", "NA", "#DIV/0!"))
```

Columns with little or no variability won't be good predictors and will be removed.  

The data sets contain many columns with no data and the first 7 columns do provide any useful infomration in our analysis.  The data will be cleaned by removing the first 7 columns and any columns with no (or NA) data.

```{r}
# remove columns with no data (and near-zero??)
#training = nearZeroVar(training, saveMetrics = TRUE)
#testing = nearZeroVar(testing, saveMetrics = TRUE)
training = training[,colSums(is.na(training)) == 0]
testing = testing[,colSums(is.na(testing)) == 0]

#write.csv(training, "pml-training-pass1.csv")
#write.csv(testing, "pml-testing-pass1.csv")

# the first 7 columns aren't required for our analysis
training = training[, -c(1:7)]
testing = testing[, -c(1:7)]

#write.csv(training, "pml-training-pass2.csv")
#write.csv(testing, "pml-testing-pass2.csv")

training$classe = as.factor(training$classe)
```


badColumns <- nearZeroVar(training, saveMetrics = TRUE)
training <- training[, badColumns$nzv==FALSE]
training$classe = factor(training$classe)



Now the data has been loaded and in a useable state, the training data set will be split into a training (80%) and validation (20%) data set for the purpose of cross-validation.

```{r}
library(caret)
set.seed(843916)
inTrain <- createDataPartition(training$classe, p = 0.6, list = FALSE)
validation <- training[-inTrain, ]
training <- training[inTrain, ]

# while testing!!
good_training = training
good_testing = testing
good_validation = validation
```


Pre processing

```{r}
preObj <- preProcess(x = training, method = c("center", "scale", "pca"), thresh = 0.9)
print(preObj)
pTrain <- predict(preObj, training)
```



```{r}
tc = trainControl(method = "cv", number = 5)

# build model
mRp = train(classe ~ ., method="rpart", data=training) # classification tree
mRf = train(classe ~ ., method="rf", data=training) # random forest (slow....)
mGbm = train(classe ~ ., method="gbm", data=training, verbose=FALSE) # boosted trees
mLda = train(classe ~ ., method="lda", data=training) # linear discriminant analysis

#library(rattle)
#fancyRpartPlot(mRp$finalModel)

# Calculation of variable importance for regression and classification models
varImpObj <- varImp(mRf)
plot(varImpObj, main = "Variables Importance", top = 30)

# predict
pRp = predict(mRp, validation)
pRf = predict(mRf, validation)
pGbm = predict(mGbm, validation)
pLda = predict(mLda, validation)

# combine the predictions
dfComb = data.frame(pRp, pRf, pGbm, pLda, classe = testing$classe)
mComb <- train(classe ~ ., dfComb)
pComb <- predict(mComb, testing)

# confusion matrix (for accuracy)
cmRp <- confusionMatrix(pRp, testing$classe)
cmRf <- confusionMatrix(pRf, testing$classe)
cmGbm <- confusionMatrix(pGbm, testing$classe)
cmLda <- confusionMatrix(pLda, testing$classe)
cmComb <- confusionMatrix(pComb, testing$classe)

# accuracy of individual and combined
cmRp$overall['Accuracy']
cmRf$overall['Accuracy']
cmGbm$overall['Accuracy']
cmLda$overall['Accuracy']
cmComb$overall['Accuracy']
```
