
# Overview

One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

# Data

The training data for this project are available here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

```{r}
# get the training and test data
if (!file.exists("pml-training.csv"))
{
    download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv",destfile="pml-training.csv",method="libcurl")
}
if (!file.exists("pml-testing.csv"))
{
    download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv",destfile="pml-testing.csv",method="libcurl")
}

# load the training and test data sets
training = read.csv("pml-training.csv", na.strings = c("", "NA", "#DIV/0!"))
testing = read.csv("pml-testing.csv", na.strings = c("", "NA", "#DIV/0!"))
```

# Objectives

You should create a report describing how you built your model

How you used cross validation

What you think the expected out of sample error is

Why you made the choices you did

You will also use your prediction model to predict 20 different test cases

Please constrain the text of the writeup to < 2000 words and the number of figures to be less than 5

link to a Github repo with your R markdown and compiled HTML file describing your analysis

# Exploratory Data Analysis

```{r}
library(caret)
set.seed(62433)

# k-means (break training into segments)

# build model - every row has at least one missing value
mRf = train(classe ~ ., method="rf", data=training)
mGbm = train(classe ~ ., method="gbm", data=training, verbose=FALSE)
mLda = train(classe ~ ., method="lda", data=training)

# predict
pRf = predict(mRf, testing)
pGbm = predict(mGbm, testing)
pLda = predict(mLda, testing)

# combine the predictions
dfComb = data.frame(pRf, pGbm, pLda, classe = testing$classe)
mComb <- train(classe ~ ., dfComb)
pComb <- predict(mComb, testing)

# confusion matrix (for accuracy)
cmRf <- confusionMatrix(pRf, testing$classe)
cmGbm <- confusionMatrix(pGbm, testing$classe)
cmLda <- confusionMatrix(pLda, testing$classe)
cmComb <- confusionMatrix(pComb, testing$classe)

# accuracy of individual and combined
cmRf$overall['Accuracy']
cmGbm$overall['Accuracy']
cmLda$overall['Accuracy']
cmComb$overall['Accuracy']
```


